---
title: "Analyze Incidents & Flights"
author: "SioWa Luo"
date: "Summer 2, 2025"
output:
  pdf_document: default
  html_document: default
subtitle: CS3200 Final Project
---

```{r connectToDatabase, eval = T, echo = F, message = F, warning = F}
# Import Necessary Library
library(RMySQL)
library(DBI)
library(kableExtra)

# Define Setting
db_host_aiven <- "mysql-cs3200-northeastern-cs3200.k.aivencloud.com"
db_port_aiven <- 22967
db_name_aiven <- "defaultdb"
db_user_aiven <- "avnadmin"
db_pwd_aiven <- "AVNS_b8LFfIb2ijYwEuui_FU"

# Embedded SSL certificate
db_cert <- 
"
-----BEGIN CERTIFICATE-----
MIIEUDCCArigAwIBAgIUDSsZAl6kvMb0yrAdHFvhOPbcJ70wDQYJKoZIhvcNAQEM
BQAwQDE+MDwGA1UEAww1YzNiYTJlNTMtNDA5NS00YjM3LWEwYmUtNDllODgwODBl
ODI4IEdFTiAxIFByb2plY3QgQ0EwHhcNMjUwODAzMjE0NDEzWhcNMzUwODAxMjE0
NDEzWjBAMT4wPAYDVQQDDDVjM2JhMmU1My00MDk1LTRiMzctYTBiZS00OWU4ODA4
MGU4MjggR0VOIDEgUHJvamVjdCBDQTCCAaIwDQYJKoZIhvcNAQEBBQADggGPADCC
AYoCggGBAL5qvx5moiDVHp44nyhjUIeYz9naW7pVPqb9ZSF+USu2IXh4zjFH24Js
hxCu/kdALiQfcgjWUisBnPmguuV+GCTqEYrmfIpD7tEmp5Zls1s4eWL4YtNYMO9p
mhHaR/mKFA9+AmDDLB13sO2+IgJxvEjDXkdDHbTnHwPgX1pUBo+d0evTZhcoqpRz
UiqvsT2pyUb8FsRH83DReyM4qaV8EzOEjM8wIZEvO29Wn6cxcRxADE7Cn1I6k3Q9
rheEYNIzU3d+vrwdQQg+fc7qTmYr6swnU7fNOGawHufkHjaXiZJZk0jztVnp/m/G
3BfenTy+UWEa3CIk5l80LRJ/QHO+kCXnlunkW5ky7DkRulcXcT30R1qF5OzQQYC1
gL4lHrd9HG6yR++3CWzwt9AxSrmbzDL2Y56wZjtJcKfNiSeHIIWZ7y1gnFHzGyJT
QTDXvETAVdNIrLbjuRWrjp11l1x4w6TgWsKoi/1sY7Wr4r+oga+UNLiQhpJJy94J
AroefkBNZwIDAQABo0IwQDAdBgNVHQ4EFgQU7X6M21DSCxWu/lDNUDII1del8YIw
EgYDVR0TAQH/BAgwBgEB/wIBADALBgNVHQ8EBAMCAQYwDQYJKoZIhvcNAQEMBQAD
ggGBAL2gxg4vYo4MVQPujzOTmrAKuRlG0uT3gwgKfyCTNbPrlQE/ux0aYdg7ZcQg
LBjHqxTJMsq7Lo/Y1c8DvqtBEcSFTfJ3A/HRJj9I5Wkuj9AzGdXX/J9RH9xdzIEU
bqxap847vC8ddrBHGD7E5K6tilRzsZe412I+ymVPylM0mDJm9YOihojfhqlBGDYR
9qDYFWcbnwsHrna/qIywLtZrJMs4xOX+Oz8LPdqGZ1+TPEahEUl6WxH0ivggorFw
lx69r+wQN6x/ETrbXLj5AkwUo0LcQi/RXN5Vis9VQcVtWUrz6Tt96k0q8x1WsR7M
93lnSusJEOn7d8NhuQ7iZdGv4qy7h1J92iWVxswYvAV15AmOOzmuFlQxBfkviFH/
KCq03oic8tsZadXAtapY4MHo6dqtHQw1Zc1e27+X4p2qOzwBK94UQlhUeTCrALXq
187ebDb5S0ZRCvxBTgzThxgfgg00jcjNr+wxkQBzf5xPkpy4nQ9qtEndZ4zJOTgh
Ac412g==
-----END CERTIFICATE-----
"

# connect to remote MySQL server and database
mydb.aiven <-  dbConnect(RMySQL::MySQL(), 
                         user = db_user_aiven, 
                         password = db_pwd_aiven,
                         dbname = db_name_aiven, 
                         host = db_host_aiven, 
                         port = db_port_aiven,
                         sslmode = "require",
                         sslcert = db_cert)
```

## Analysis by Month
The following will be the table contain the total number of incident by type and month:
```{r monthAnalysisTable, eval = T, echo = F, message = F, warning = F}
# Query to get relevant data for month and count of each incident type
monthly_incident_query <- paste0(
  "SELECT ",
  "MONTHNAME(i.incidentDate) as Month, ",
  "it.incidentTypeName as IncidentType, ",
  "COUNT(*) as TotalIncidents ",
  "FROM Incidents i ",
  "JOIN IncidentTypes it ON i.incidentTypeId = it.incidentTypeId ",
  "GROUP BY MONTH(i.incidentDate), MONTHNAME(i.incidentDate), it.incidentTypeName ",
  "ORDER BY MONTH(i.incidentDate), it.incidentTypeName"
)

# Execute the query
monthly_incident_data <- dbGetQuery(mydb.aiven, monthly_incident_query)

# Getting unique month and incident type 
unique_months <- unique(monthly_incident_data$Month)  
unique_types <- unique(monthly_incident_data$IncidentType) 

# Initiate empty df for month and type set to 0
incidents_df <- data.frame(Month = unique_months)
for(type in unique_types) {
  incidents_df[[type]] <- 0 
}

# Fill df with incident counts 
# Loop through each row and get month, incident type and calculate the counts
for(i in 1:nrow(monthly_incident_data)) {
  month <- monthly_incident_data$Month[i]     
  type <- monthly_incident_data$IncidentType[i] 
  count <- monthly_incident_data$TotalIncidents[i]  
  
  # Find the row for specific month and set the count the expected incident type
  row_index <- which(incidents_df$Month == month)
  incidents_df[row_index, type] <- count
}

# Add col to calculate sum accross the incident type
incidents_df$Total <- rowSums(incidents_df[, -1])

# Create formatted table using kable
kable(incidents_df, 
      caption = "Total Number of Incidents by Type and Month",
      align = "c")
```
```{r monthAnalysisStats, eval = T, echo = F, message = F, warning = F}
# Calculate monthly stats
monthly_totals <- rowSums(incidents_df[, -c(1, ncol(incidents_df))])
min_month <- incidents_df$Month[which.min(incidents_df$Total)]
max_month <- incidents_df$Month[which.max(incidents_df$Total)]
min_total <- min(incidents_df$Total)
max_total <- max(incidents_df$Total)

summer_months <- c("May", "June", "July", "August", "September")
summer_avg <- mean(incidents_df$Total[incidents_df$Month %in% summer_months])

winter_months <- c("December", "January", "February")
winter_avg <- mean(incidents_df$Total[incidents_df$Month %in% winter_months])
```

The monthly incident table shows a strong seasonal pattern in bird strike activity. Winter months seems to have a lower activity with `r min_month` having the least incidents (`r min_total` incidents) and `r max_month` show most acivity with `r max_total` incidents. In Summer month (May - September), the average `r round(summer_avg, 0)` incidents per month while the Winter month (December - February) average only `r round(winter_avg, 0)` incidents per month. This is around `r round((summer_avg - winter_avg) / winter_avg * 100, 0)` % increased during the warmer months. The reason to this might due to the bird migration season which makes summer months a more critical periods for preventing bird strike incident.

## Analysis by Airline
```{r airlineAnalysis, eval = T, echo = F, message = F, warning = F}
# Query to create a view
# The view summarizes incident data by airline and calculates summary statistics
create_view_query <- paste0(
  "CREATE OR REPLACE VIEW AirlineIncidentSummary AS ",
  "SELECT ",
  "a.airline_name as Airline, ",
  "COUNT(i.iid) as TotalIncidents, ",
  "ROUND(AVG(CASE WHEN i.delayMinute > 0 THEN i.delayMinute ELSE NULL END), 2) as AverageDelay ",
  "FROM Airlines a ",
  "JOIN Flights f ON a.airlineId = f.airlineId ",
  "JOIN Incidents i ON f.flightId = i.flightId ",
  "GROUP BY a.airlineId, a.airline_name ",
  "ORDER BY TotalIncidents DESC"
)

# Execute the query to create the view
invisible(dbExecute(mydb.aiven, create_view_query))

# Query to get all data from view we created
airline_summary_query <- paste0(
  "SELECT * FROM AirlineIncidentSummary"
)

# Execute the query
airline_summary <- dbGetQuery(mydb.aiven, airline_summary_query)

# Create formatted table for airline summary
airline_summary_col <- c("Airline", "Total Incidents", "Average Delay (minutes)")

# Create table for airline incident summary
kable(airline_summary,
      caption = "Incident Summary and Average Delay by Airline",
      col.names = c("Airline", "Total Incidents", "Average Delay (minutes)"),
      align = c("l", "c", "c"))
```

```{r airlineAnalysisStats, eval = T, echo = F, message = F, warning = F}
# Calculate airline stats
top_airline <- airline_summary$Airline[1]
top_incidents <- airline_summary$TotalIncidents[1]

bottom_airline <- airline_summary$Airline[nrow(airline_summary)]
bottom_incidents <- airline_summary$TotalIncidents[nrow(airline_summary)]

incident_range <- top_incidents - bottom_incidents

avg_delay_min <- min(airline_summary$AverageDelay)
avg_delay_max <- max(airline_summary$AverageDelay)

best_delay_airline <- airline_summary$Airline[which.min(airline_summary$AverageDelay)]
worst_delay_airline <- airline_summary$Airline[which.max(airline_summary$AverageDelay)]

delay_variance <- avg_delay_max - avg_delay_min
```

The airline analysis shows the performance between airline, with total incidents  from `r bottom_incidents` (`r bottom_airline`) to `r top_incidents` (`r top_airline`) with a difference of `r incident_range` incidents. This show us that the incident is mostly correlates with operation volume. We observe a average delay vary by `r round(delay_variance, 1)` minutes between the best (`r best_delay_airline` at `r avg_delay_min` minutes) and worst (`r worst_delay_airline` at `r avg_delay_max` minutes) airline. The small variance shows overall industry protocols for incident response. The differences are likely due to other operational reason than the safety protocol issue.

## Trend by Year
```{r plotTrendByYear, eval = T, echo = F, message = F, warning = F}
# Create SQL query using paste0
yearly_incidents_query <- paste0(
  "SELECT ",
  "YEAR(incidentDate) as Year, ",
  "COUNT(*) as TotalIncidents ",
  "FROM Incidents ",
  "GROUP BY YEAR(incidentDate) ",
  "ORDER BY Year"
)

# Execute the query
yearly_incident_data <- dbGetQuery(mydb.aiven, yearly_incidents_query)

# Get data for x and y axis
x <- yearly_incident_data$Year
y <- yearly_incident_data$TotalIncidents

# Create line chart
# Setting up the plot
plot(x, y, 
     type = "o", pch = 16, col = "steelblue",
     main = "Bird Strike Incidents Trend by Year",
     xlab = "Year", ylab = "Total Number of Incidents",
     ylim = c(500, 1250))


# Add data labels on top of each data (total incidents by year)
# Reference https://www.spsanderson.com/steveondata/posts/2023-08-07/index.html
text(x, y, labels = yearly_incident_data$TotalIncidents,
     pos = 3, cex = 0.8, col = "black")


# Add legend
par(xpd = TRUE)
legend(x = 2022,y = 1450,
       legend = c("Incidents"),
       col = c("steelblue"),
       lwd = 2,
       pch = c(16, NA))
```

```{r trendByYearStats, eval = T, echo = F, message = F, warning = F}
# Calculate yearly stats
total_years <- nrow(yearly_incident_data)

start_year <- min(yearly_incident_data$Year)
end_year <- max(yearly_incident_data$Year)

start_incidents <- yearly_incident_data$TotalIncidents[1]
end_incidents <- yearly_incident_data$TotalIncidents[nrow(yearly_incident_data)]

# Calculate data for raising trend and stable period
raising_incidents <- yearly_incident_data$TotalIncidents[2]
stable_start_year <- 2009
stable_end_year <- 2022
stable_period <- yearly_incident_data[yearly_incident_data$Year >= stable_start_year & yearly_incident_data$Year <= stable_end_year, ]
stable_min <- min(stable_period$TotalIncidents)
stable_max <- max(stable_period$TotalIncidents)
avg_incidents <- round(mean(stable_period$TotalIncidents), 1)

# Calculate peak
peak_year <- yearly_incident_data$Year[which.max(yearly_incident_data$TotalIncidents)]
peak_incidents <- max(yearly_incident_data$TotalIncidents)

# Disconnect from DB
invisible(dbDisconnect(mydb.aiven))
```
The visualization analayze `r total_years` year from `r start_year` to `r end_year` showing the trend for airline safety evlotion. It start with `r start_incidents` in `r start_year`, but in the next year the incidents raise dramatically to `r raising_incidents` incidents in `r stable_start_year` and then maintain this level between `r stable_min` - `r stable_max` incidents annually through  `r stable_end_year` with average of `r avg_incidents` incidents per year. In `r peak_year` there is a peak occur with `r peak_incidents` incidents. And at the very end of the trend, there is a sharp decline to `r end_incidents` incidents in `r end_year` which create a situation that the beginning and ending years are having a significantly lower incidents acitivities compare to the middle of the trend. This situation shows that there there are some factor that might cause the significant changes during early and end of the periods. Some posibility might be in the early stage there is a new prevention system is being added to the airline operation and it is taking sometimes for them to adapt to the system in the middle which cause the higher number of incidents and in the end new updates is made into the system and makes it more effective which reduce the incidents activities.



